### 2022-08-25 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex

@@

\narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
  facet model's sensitivity to small data perturbations. This example is
  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have the same
  curvature (equal second derivatives). Given the exact five data points
  seen above, the quadratic facet model produces the slope seen in the
  solid blue line at $x = 5/2$. However, by subtracting the value of
  $f_3$ $= \epsilon(x-2)^2$ from points at $x = 3$, 4, where $\epsilon$
  is the machine precision ($2^{-52}$ for an IEEE 64-bit real), the
  quadratic facet model produces the slope seen in the dashed red line
  at $x = 5/2$. This is the nature of a facet model and a side effect
  of associating data with local facets.
\par} \endinsert

>>>>

\narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
  facet model's sensitivity to small data perturbations. This example is
  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have
  \red{equal second derivatives}. Given the exact five data points
  seen above, the quadratic facet model produces the slope seen in the
  solid blue line at $x = 5/2$. However, by subtracting the value of
  $f_3$ $= \epsilon(x-2)^2$ from points at $x = 3$, 4, where $\epsilon$
  is the machine precision ($2^{-52}$ for an IEEE 64-bit real), the
  quadratic facet model produces the slope seen in the dashed red line
  at $x = 5/2$. This is the nature of a facet model and a side effect
  of associating data with local facets.
\par} \endinsert

@@

The binary search procedure provably converges on the boundary of the
region of monotonicity precisely defined by Ulrich and Watson [1994]
through the application of a Boolean function that is guaranteed to be
true at one end of an interval. Since a local quadratic interpolant is
applied for initial function derivative estimates, the approximation
order of the resulting fit is $O(h^3)$ (as for any second order
approximation), but it should be noted that asymptotic approximation
order is rarely of importance when considering the scattered sparse
data that $C^2$ approximations like {\tt MQSI} provide. Were exact
evaluations provided, higher order methods always win (all
other things being equal) as the data density increases. For sparse
data, what constitutes a better fit is either subjective or dependent
on the problem.

### 2022-08-25 [2020-08]_ACMTOMS_(MQSI)/paper/appendix.tex

\vfil\break

@@

The quiet failure of two published codes for shape
preserving spline interpolation is surprising, and accentuates how
difficult it is to correctly handle the numerical intricacies of
polynomial function evaluation. For the sake of reproducibility, coded
minimum working examples of the failure modes for the Schumaker spline
package in {\tt R} and the {\tt BVSPIP} package in Fortran are
provided.

\vskip 5mm

\noindent{\bf Schumaker error in {\tt R}}

\vskip 3mm

{\parskip=-1pt \parindent=10pt \ttVIII
\item{} \textMaroon \# Load the library and set up the problem data. \textBlack
\item{} library(schumaker)
\item{} x = c(0.0, 0.001, 0.02, 0.03)
\item{} y = c(0.0, 0.09, 0.1, 0.11)
\item{} \textMaroon \# Construct the spline fit, rename its evaluator as {\char13}f{\char13}. \textBlack
\item{} SchumakerSpline = schumaker::Schumaker(x, y)
\item{} f = SchumakerSpline\$Spline
\item{} \textMaroon \# Evaluate at 0.01 to get 0.28, greater than all {\char13}y{\char13}. \textBlack
\item{} f(0.01) > max(y)
\item{} \textMaroon \# Plot the results to see the failed approximation. \textBlack
\item{} plot(x, y, col=1, ylim=c(-0.1,0.35), main="Schumaker", ylab="y", xlab="x")
\item{} lines(seq(0,0.03,0.0001), f(seq(0,0.03,0.0001)), col=4)
}

\vskip 5mm

\noindent{\bf BVSPIS error in Fortran}

\vskip 3mm

{\parskip=0pt \parindent=10pt \ttVIII
\item{} \textMaroon ! Compile and execute the following program with: \textBlack
\item{} \textMaroon ! \$F03 bvspis.f \$FILENAME -o \$EXECNAME \textBlack
\item{} \textMaroon ! ./\$EXECNAME \textBlack
\item{} \textMaroon ! Where some placeholder values for the above shell variables might result in: \textBlack
\item{} \textMaroon ! gfortran bvspis.f test\_bvspis\_errors.f03 -o test\_bvspis\_errors \textBlack
\item{} \textMaroon ! ./test\_bvspis\_errors \textBlack
\item{} \textCyan PROGRAM \textBlue TEST\_BVSPIS \textBlack
\item{}   \textCyan IMPLICIT \textGreen NONE\textBlack
\item{}   \textCyan EXTERNAL \textBlack :: \textBlue DBVSSC\textBlack, \textBlue DBVSSE\textBlack
\item{}   \textGreen INTEGER \textBlack :: NP, N, K, R, Q, P, OPT, KMAX, MAXSTP, NWORK, SBOPT, NTAB, \textCyan\&\textBlack $ $
\item{}   \hskip 10mm Y0OPT, Y1OPT, Y2OPT, ERRC, ERRE
\item{}   \textGreen DOUBLE PRECISION \textBlack :: EPS, D0, DNP, D20, D2NP, BETA, BETAI, RHO, RHOI
\item{}   \textGreen INTEGER\textBlack, \textCyan ALLOCATABLE \textBlack :: CONSTR(:), DIAGN(:) 
\item{}   \textGreen DOUBLE PRECISION\textBlack, \textCyan ALLOCATABLE \textBlack :: X(:), Y(:), XTAB(:), D(:), D2(:), \textCyan\&\textBlack $ $
\item{}   \hskip 10mm WORK(:), Y0TAB(:), Y1TAB(:), Y2TAB(:)
\item{}   \textMaroon ! Set parameters. \textBlack
\item{}   EPS = \textCyan SQRT\textBlack(\textCyan EPSILON\textBlack(0.0D0)) \textMaroon ! Default precision. \textBlack
\item{}   NP = 3-1 \textMaroon ! Number of points minus one. \textBlack
\item{}   N = 6 \textMaroon ! Degree of the spline (minimum of 3*k). \textBlack
\item{}   K = 2 \textMaroon ! Class of continuity (C2). \textBlack
\item{}   R = 1 \textMaroon ! Enforce monotonicity constraints. \textBlack
\item{}   Q = 1 \textMaroon ! No boundary conditions are imposed. \textBlack
\item{}   P = 1 \textMaroon ! Derivative estimation by constraints only. \textBlack
\item{}   OPT = 100*P + 10*Q + R \textMaroon ! Contains PQR in decimal format. \textBlack
\item{}   NWORK = 5+(2+7)*NP+(N*(N+11))/2+9 \textMaroon ! Explained in DBVSSC comments. \textBlack
\item{}   SBOPT = 2 \textMaroon ! Do a binary search to find intervals for evaluation. \textBlack
\item{}   NTAB = 3-1 \textMaroon ! Number of evaluation points minus one. \textBlack
\item{}   Y0OPT = 1 \textMaroon ! Enable evaluation of the zeroth derivative. \textBlack
\item{}   Y1OPT = 0 \textMaroon ! Disable evaluation of the first derivative. \textBlack
\item{}   Y2OPT = 0 \textMaroon ! Disable evaluation of the second derivative. \textBlack
\item{}   ERRC = 0 \textMaroon ! Error code, initialize to 0, should be overwritten. \textBlack
\item{}   ERRE = 0 \textMaroon ! Error code, Initialize to 0, should be overwritten. \textBlack
\item{}   \textMaroon ! Allocate and initialize arrays. \textBlack
\item{}   \textCyan ALLOCATE\textBlack( \textCyan\&\textBlack $ $
\item{}   \hskip 10mm X(0:NP), \textCyan\& \textBlack \textMaroon ! X values that cause failure. \textBlack $ $
\item{}   \hskip 10mm Y(0:NP), \textCyan\& \textBlack \textMaroon ! Y values that cause failure. \textBlack $ $
\item{}   \hskip 10mm CONSTR(0:NP), \textCyan\& \textBlack \textMaroon ! Used for setting shape preserving constraints. \textBlack $ $
\item{}   \hskip 10mm XTAB(0:NTAB), \textCyan\& \textBlack \textMaroon ! Z evaluation point (known failure location). \textBlack $ $
\item{}   \hskip 10mm D(0:NP), \textCyan\& \textBlack \textMaroon ! Derivative values (on output). \textBlack $ $
\item{}   \hskip 10mm D2(0:NP), \textCyan\& \textBlack \textMaroon ! Referenced with K=2. \textBlack $ $
\item{}   \hskip 10mm DIAGN(0:NP-1), \textCyan\& \textBlack \textMaroon ! Diagnostic info. \textBlack $ $
\item{}   \hskip 10mm WORK(1:NWORK), \textCyan\& \textBlack \textMaroon ! Work space. \textBlack $ $
\item{}   \hskip 10mm Y0TAB(0:NTAB), \textCyan\& \textBlack \textMaroon ! Spline values. \textBlack $ $
\item{}   \hskip 10mm Y1TAB(0:NTAB), \textCyan\& \textBlack \textMaroon ! First derivative values. \textBlack $ $
\item{}   \hskip 10mm Y2TAB(0:NTAB) \textCyan\& \textBlack \textMaroon ! Second derivative values. \textBlack $ $
\item{}   )
\item{}   X(0:NP) = (/ 0.0D0, 0.09D0, 0.091D0 /)
\item{}   Y(0:NP) = (/ 0.0D0, 0.1D0, 0.2D0 /)
\item{}   XTAB(0:NTAB) = (/ 0.0D0, 0.06D0, 0.09D0 /)
\item{}   CONSTR(0:NP) = 1
\item{}   D(0:NP) = 0.0D0
\item{}   D2(0:NP) = 0.0D0
\item{}   DIAGN(0:NP-1) = 0.0D0
\item{}   WORK(1:NWORK) = 0.0D0
\item{}   Y0TAB(0:NTAB) = 0.0D0
\item{}   Y1TAB(0:NTAB) = 0.0D0
\item{}   Y2TAB(0:NTAB) = 0.0D0
\item{}   \textMaroon ! Unused parameters, ignored unless Q=3. \textBlack
\item{}   D0 = 0.0D0
\item{}   DNP = 0.0D0
\item{}   D20 = 0.0D0
\item{}   D2NP = 0.0D0
\item{}   \textMaroon ! Unused parameters, ignored unless Q=2. \textBlack
\item{}   BETA = 0.0D0
\item{}   BETAI = 0.0D0
\item{}   RHO = 0.0D0
\item{}   RHOI = 0.0D0
\item{}   KMAX = 0
\item{}   MAXSTP = 0
\item{}   \textMaroon ! Construct the fit. \textBlack
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}Constructing fit.{\char13}\textBlack
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}X(:){\char13}\textBlack, X(:)
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}Y(:){\char13}\textBlack, Y(:)
\item{}   \textCyan CALL \textBlue DBVSSC\textBlack( \textCyan\&\textBlack $ $
\item{}   \hskip 10mm X(:), Y(:), NP, N, K, OPT, D0, DNP, D20, D2NP, CONSTR(:), \textCyan\&\textBlack $ $
\item{}   \hskip 10mm EPS, BETA, BETAI, RHO, RHOI, KMAX, MAXSTP, ERRC, \textCyan\&\textBlack $ $
\item{}   \hskip 10mm D(:), D2(:), DIAGN(:), WORK(:), NWORK \textCyan\&\textBlack $ $
\item{}   )
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}ERRC {\char13}\textBlack, ERRC \textMaroon ! = 0 \textBlack
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}{\char13}\textBlack
\item{}   \textMaroon ! Evaluate the fit. \textBlack
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}Evaluate the spline.{\char13}\textBlack
\item{}   \textCyan CALL \textBlue DBVSSE\textBlack( \textCyan\&\textBlack $ $
\item{}   \hskip 10mm X(:), Y(:), NP, N, K, XTAB(:), NTAB, SBOPT, Y0OPT, Y1OPT, \textCyan\&\textBlack $ $
\item{}   \hskip 10mm Y2OPT, ERRC, D(:), D2(:), Y0TAB(:), Y1TAB(:), Y2TAB(:), \textCyan\&\textBlack $ $
\item{}   \hskip 10mm ERRE, WORK(:), NWORK \textCyan\&\textBlack $ $
\item{}   )
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}XTAB(:) {\char13}\textBlack, XTAB(:) \textMaroon ! = 0.0, 0.06, 0.09 as expected.\textBlack
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}Y0TAB(:){\char13}\textBlack, Y0TAB(:) \textMaroon ! = 0.0, 0.152, 0.1 not monotone, unexpected.\textBlack
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}ERRE {\char13}\textBlack, ERRE \textMaroon ! = 0 no error code, unexpected.\textBlack
\item{}   \textCyan PRINT \textBlack *, \textRed{\char13}ERRC {\char13}\textBlack, ERRC \textMaroon ! = 0 no error code, unexpected.\textBlack
\item{} \textCyan END PROGRAM \textBlue TEST\_BVSPIS \textBlack

}

>>>>

\beginred

@@

The quiet failure of two published codes for shape preserving spline
interpolation is surprising, and accentuates how difficult it is to
correctly handle the numerical intricacies of polynomial function
evaluation. For the sake of reproducibility, coded minimum working
examples of the failure modes for the Schumaker spline package in {\tt
  R} and the {\tt BVSPIP} package in Fortran are provided as
supplemental files. The program {\tt schumaker\_test.r} demonstrates a
sequence of four monotone increasing points that yields a nonmonotone
approximation. The program {\tt bvspis\_test.f03} provides a sequence
of three monotone increasing points for which {\tt BVSPIS} produces a
nonmonotone approximation.

\endred
